---
title: "class12"
author: "Rosa Smith"
format: pdf
editor: visual
---

## 

```{r}
counts <-read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

```{r}
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

```{r}
nrow(counts)
nrow(metadata)
```

## Q1. Based on data above, there are 38694 genes in the dataset.

```{r}
View(metadata)
```

##Q2. After viewing the metadata, there are 4 control cell lines.

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```
I like the dplyr package because it makes sense in my head to "filter" for different variables.

## Q3. You could assign the 4 values to a vector that is the actual number of controls in case the value were to change.

##Q4. shown below.
```{r}
library(dplyr)

treated <- metadata %>% filter(dex == "treated")

treated.counts <- counts %>% select(treated$id)

treated.mean <- rowMeans(treated.counts)

head(treated.mean)
```
```{r}
#combining meancount data

meancounts <- data.frame(control.mean, treated.mean)
```


## Q5a. shown below.

```{r}
plot(meancounts)
```

```{r}
library(ggplot2)
ggplot(meancounts, aes(control.mean, treated.mean))+
  geom_point(alpha=.4)
#alpha used to change opacity
```
## Q5b. shown above. Using "geom_point" in the ggplot2 package.
```{r}
library(ggplot2)
ggplot(meancounts, aes(control.mean, treated.mean))+
  geom_point(alpha=.4)+
  scale_x_log10()+
  scale_y_log10()
  
#scaled using ggplot

```

```{r}
plot(meancounts, log="xy")
```
## Q6. shown above with ggplot and default plot().

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
#filtered data
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

## Q7. arr.ind = TRUE makes which() return the row and column positions of all TRUE values instead of just a single vector of positions, and gives rows and columns that have zeroes. Then using unique() so that if it a gene has zeros in multiple samples it's only listed once.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```
```{r}
table(up.ind)
```
```{r}
table(down.ind)
```


## Q8. Using results above, there are 250 up regulated changes at the greater than 2 fc level.

## Q9. From the table above, there are 367 down regulated genese at the greater than 2 fc level.

## Q10. The results could be misleading because it's based on fold-changes, which could be large without being statistically significant.


```{r}
library(DESeq2)
citation("DESeq2")

```
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```
```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

```{r}
dds <- DESeq(dds)
```

```{r}
# results

res <- results(dds)
res

```
```{r}
summary(res)
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

```
```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # genenames
                     keytype="ENSEMBL",        #format of our genenames
                     column="SYMBOL",          # The new format 
                     multiVals="first")

head(res)
```

## Q11. Shown below.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)

```
```{r}
# viewing by p-value

ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```
```{r}
write.csv(res[ord,], "deseq_results.csv")
```
```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")

```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```
```{r}

mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```
```{r}


library(EnhancedVolcano)

x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```


